Visual Diagnostics of a Model Explainer -- Tools for the Assessment of LIME Explanations

### 2020-08-27 paper.Rnw
changed a few past tense words to present tense

\kge{Before introducing the visual diagnostics, let us consider a commonly used visualization of a LIME explanation that is shown in Figure \ref{fig:sine-exp-scatter} (left). The plot is created using the \emph{lime} R package and provides a visual representation of a LIME explanation for the prediction of interest indicated in \autoref{fig:sine-plot-data}. In particular, the explanation was obtained using 4 quantile bins to simulate the data.} \kgc{I moved the next few sentences here from the sub-section.} In this scenario, LIME converts continuous predictor variables to indicator variables identifying whether the variable value falls in the same quantile bin as the prediction of interest or not. The indicator variables are used as the explainer model features. \sout{To demonstrate a LIME explanation, we focus on the misclassified point indicated in Figure \ref{fig:sine-plot-data} by the diamond as the prediction of interest. The left plot in Figure \ref{fig:sine-exp-scatter} shows a visualization of the explanation for the prediction of interest as suggested by Ribeiro et al and implemented in the \emph{lime} R package.}

@@

The explanation for the prediction of interest depicted in Figure \ref{fig:sine-exp-scatter} can be interpreted as follows. A random forest classification of blue is supported by the prediction of interest having a value of $x_2$ that is greater than \Sexpr{str_split(sine_poi_lime_default$feature_desc[[1]], pattern = " ")[[1]][1]}, but the prediction of interest having a value of $x_1$ that is greater than \Sexpr{str_split(sine_poi_lime_default$feature_desc[[2]], pattern = " ")[[1]][1]} and less than or equal to \Sexpr{str_split(sine_poi_lime_default$feature_desc[[2]], pattern = " ")[[1]][5]} provides support against a classification of blue. \kge{Since the weight associated with $x_2$ has a larger magnitude than the weight associated with $x_1$, LIME explains that $x_2$ plays a more important role in the random forest prediction.} \kge{Additionally, consider that since the support for blue by $x_2$ outweights the support for red by $x_1$, the explainer model overall favors a label of blue for the prediction of interest, which} agrees with the random forest probability of \Sexpr{sine_poi_lime_default$label_prob[1]} for blue. Note that both models classify the observation incorrectly.

@@

First, consider the claim of localness. The weights \kge{decay relatively slowly} \sout{remain relatively high} outside of the intersection of the two quantile bins suggesting that the LIME explanation is highly influenced by points outside of the bin\kge{s} containing the prediction of interest. \sout{While i} \kge{However,} it is \sout{still} difficult to say \kge{if} \sout{whether or not} the claim of localness has been violated\kge{, because a definition of a local region was not specified. Depending a the region a viewer considers to be local, an argument could be made in favor of or against a violation of localness. The explanation scatterplot raises awareness of the unclear definition of a local region with LIME. Nevertheless,} it is possible to say that the weights assigned to the simulated data do not \kge{capture} \sout{agree with} the local region \kge{identified} \sout{assigned} by the intersection of the quantile bins.

@@

Two hypothetical examples of feature heatmaps are included in \autoref{fig:concept-feat-heat}. The plots were created with the assumption that LIME is applied to select the top feature out of $p=\kge{4}$ \sout{8} features for $n=10$ cases with $t=5$ sets of tuning parameters. Situation 1 \kge{is} \sout{shows} an example where the features selected are consistent across tuning parameters within a case but vary across cases within a tuning parameter. This is the ideal situation, because the LIME explanations do not depend on the tuning parameters but do depend on the location of the observation in the feature space. Situation 2 \kge{is} \sout{shows} an example where the features selected vary across tuning parameters within a case but are consistent across cases within a tuning parameter. This situation indicates that the features selected by LIME are dependent on the tuning parameters, and the explanations \kge{may not be} \sout{are not} local, because the same feature is chosen regardless of the case. In practice, it is expected that the plot will exhibit a combination of these two situations.

@@

LIME \sout{is implemented using the R package \emph{limeaid} to apply LIME} \kge{is applied} \sout{multiple times} to all test set observations using different tuning parameters: 12 sampling methods (2-6 equally spaced bins, 2-6 quantile bins, kernel density estimation, and normal approximation) and 3 Gower exponents (0.5, 1, and 10). Thus, a total of $12\times 3=36$ different applications of LIME were performed. We specify that each LIME explanation return 3 features and feature selection is performed \kge{by selecting the features with the highest weights in a ridge regression model} \sout{using the highest weights method} (default option).

@@

% Do I need to include this?
% \section*{Author Biography}
% \begin{biography}
% {\includegraphics[width=60pt,height=70pt,draft]{empty}}
% {\textbf{Author Name.} This is sample author biography text this is sample author biography text}
% \end{biography}

% \textsc{\textbf{Ideas for describing proposed plots}}
% 
% \begin{itemize}
% \item start at the high level explanation (this is what we want to do and this is how we do it)
% \item go over a good and bad example
% \item include schematics of the plot first (good and bad) and then include the sine mini example version
% \item explain the expectation of the plot
% \end{itemize}
% 
% \textsc{\textbf{Comments for Editing}}
% 
% \begin{itemize}
% \item \hh{Generally: avoid 'can be'. Replace by 'is'.  }
% \item \hh{References like 'they', 'them' ... replace them by repeating the noun you are referring to to avoid any kind of ambiguity}
% \item \kgc{words to go back and make sure I am not using too often: ability, produce, understand}
% \item \kgc{decide on features or predictor variables}
% \item \hh{it's tuning parameters, not tunning} \kgc{Oops! I get that one wrong all of the time}
% \item \kgc{Adjust text sizes in plots so they all match}
% \item \kgc{Color scheme - comment from Heike:} \hh{I like the red-blue scheme for binary values, I also like the color scheme in the features heatmap. Generally papers and color don't go well together anyways, but we will solve that problem when we have to ... Most likely we will have to just pay the extra color fee}
% \item \kgc{Check that I am consistently referring to the perturbing as data simulation}
% \item \kgc{Clean up notation}
% \item \kgc{Referring to the linearity assumption:} \hh{Be a bit careful in the phrasing here - we can always approximate any function using a linear form (think Taylor expansion). The question is how big the error is.}
% \item \kgc{Make sure plots in the same figure are the same size}
% \item \kgc{Make sure I am distinguishing between RF predictions and probabilities correctly}
% \item \kgc{Use prediction of interest not case of interest}
% \item \kgc{Make sure I always treat the word 'data' as plural}
% \item \kgc{dataset or data set?}
% \item \kgc{scatterplot or scatter plot (check other plot names as well)}
% \item \kgc{choose whether to use the term feature or variable and make sure it is consistent throughout}
% \item \kgc{just include the details about LIME that are necessary to understand the plots and remove the rest to the appendix}
% \item \hh{avoid the use of implicit references like `this' and `that'. Make the reference explicit by adding what you are referring to, such as `this method', `that procedure'}
% \item \hh{make sure to use exact phrases. things like `LIME was developed in 2016' is factually questionable - but we know LIME was introduced in 2016. Also `the authors were interested in ...' implies that you know their intent. It's better to not guess any intent but stick to what we know.}
% \item \hh{`our` diagnostics ... you are using `our` quite often. It is good to claim work, but it is not good to claim things that we want others to use, because that is by definition excluding everybody but you. It's better to re-phrase with a focus on what you did. We introduce, we suggest, etc.}
% \item \kgc{check that I am always using present tense}
% \item \kgc{check that I am consistent in my use of the term black-box model}
% \end{itemize}

>>>>

\kge{Before introducing the visual diagnostics, let us consider a commonly used visualization of a LIME explanation that is shown in Figure \ref{fig:sine-exp-scatter} (left). The plot is created using the \emph{lime} R package and provides a visual representation of a LIME explanation for the prediction of interest indicated in \autoref{fig:sine-plot-data}. In particular, the explanation is obtained using 4 quantile bins to simulate the data.} \kgc{I moved the next few sentences here from the sub-section.} In this scenario, LIME converts continuous predictor variables to indicator variables identifying whether the variable value falls in the same quantile bin as the prediction of interest or not. The indicator variables are used as the explainer model features. \sout{To demonstrate a LIME explanation, we focus on the misclassified point indicated in Figure \ref{fig:sine-plot-data} by the diamond as the prediction of interest. The left plot in Figure \ref{fig:sine-exp-scatter} shows a visualization of the explanation for the prediction of interest as suggested by Ribeiro et al and implemented in the \emph{lime} R package.}

@@

The explanation for the prediction of interest depicted in Figure \ref{fig:sine-exp-scatter} \kge{is} \sout{can be} interpreted as follows. A random forest classification of blue is supported by the prediction of interest having a value of $x_2$ that is greater than \Sexpr{str_split(sine_poi_lime_default$feature_desc[[1]], pattern = " ")[[1]][1]}, but the prediction of interest having a value of $x_1$ that is greater than \Sexpr{str_split(sine_poi_lime_default$feature_desc[[2]], pattern = " ")[[1]][1]} and less than or equal to \Sexpr{str_split(sine_poi_lime_default$feature_desc[[2]], pattern = " ")[[1]][5]} provides support against a classification of blue. \kge{Since the weight associated with $x_2$ has a larger magnitude than the weight associated with $x_1$, LIME explains that $x_2$ plays a more important role in the random forest prediction.} \kge{Additionally, consider that since the support for blue by $x_2$ outweights the support for red by $x_1$, the explainer model overall favors a label of blue for the prediction of interest, which} agrees with the random forest probability of \Sexpr{sine_poi_lime_default$label_prob[1]} for blue. Note that both models classify the observation incorrectly.

@@

First, consider the claim of localness. The weights \kge{decay relatively slowly} \sout{remain relatively high} outside of the intersection of the two quantile bins suggesting that the LIME explanation is highly influenced by points outside of the bin\kge{s} containing the prediction of interest. \sout{While i} \kge{However,} it is \sout{still} difficult to say \kge{if} \sout{whether or not} the claim of localness has been violated\kge{, because a definition of a local region is not specified. Depending a the region a viewer considers to be local, an argument could be made in favor of or against a violation of localness. The explanation scatterplot raises awareness of the unclear definition of a local region with LIME. Nevertheless,} it is possible to say that the weights assigned to the simulated data do not \kge{capture} \sout{agree with} the local region \kge{identified} \sout{assigned} by the intersection of the quantile bins.

@@

Two hypothetical examples of feature heatmaps are included in \autoref{fig:concept-feat-heat}. The plots \kge{are} \sout{were} created with the assumption that LIME is applied to select the top feature out of $p=\kge{4}$ \sout{8} features for $n=10$ cases with $t=5$ sets of tuning parameters. Situation 1 \kge{is} \sout{shows} an example where the features selected are consistent across tuning parameters within a case but vary across cases within a tuning parameter. This is the ideal situation, because the LIME explanations do not depend on the tuning parameters but do depend on the location of the observation in the feature space. Situation 2 \kge{is} \sout{shows} an example where the features selected vary across tuning parameters within a case but are consistent across cases within a tuning parameter. This situation indicates that the features selected by LIME are dependent on the tuning parameters, and the explanations \kge{may not be} \sout{are not} local, because the same feature is chosen regardless of the case. In practice, it is expected that the plot will exhibit a combination of these two situations.

@@

LIME \sout{is implemented using the R package \emph{limeaid} to apply LIME} \kge{is applied} \sout{multiple times} to all test set observations using different tuning parameters: 12 sampling methods (2-6 equally spaced bins, 2-6 quantile bins, kernel density estimation, and normal approximation) and 3 Gower exponents (0.5, 1, and 10). Thus, a total of $12\times 3=36$ different applications of LIME \kge{are} \sout{were} performed. We specify that each LIME explanation return 3 features and feature selection is performed \kge{by selecting the features with the highest weights in a ridge regression model} \sout{using the highest weights method} (default option).

@@
